Towards Accelerating Deep Neural Network Training on FPGAs Facilitating the Use of Variable Precision

### 2018-07-03 SWoPP2018 Paper/paper.tex
Fixed weird margin issues with the bibliography, added some text on FGPAs and the II.

  %\usepackage[dvips]{graphicx}
  \usepackage{latexsym}
  \usepackage{url}

@@

FPGAs can be used for a great variety of things, from accelerating computational algorithms to implementing a hardware design before taping out full chip.
They are devices that can represent reconfigurable hardware.

The fabric they are made up of consists of vast arrays of switches and other routing hardware to connect all the functional elements, like multipliers, block RAM and look up tables, in any way possible.
This makes them very suitable for the research described in this paper.
It gives total freedom to design a hardware implementation to explore algorithms in a hardware setting.

@@

\subsection{Neural Networks}
An Artificial Neural Network (ANNs) is a system inspired by the biological nervous system-- a system built around the concepts of \textit{neuron} as the main computational agents.
\Cref{fig:ann-example} illustrates an example of a typical “deep” neural network: a network that consists of a input layer, and output layer, and a number of \textit{hidden} layers in-between.
The number of hidden layers in modern neural networks can be very large: AlexNet (2012)~\cite{krizhevsky2012imagenet} has eight, VGG16 (2014)~\cite{simonyan2014very} has 16 layers, GoogleNet (2014)~\cite{szegedy2015going} has 22 layers, and ResNet (2016)~\cite{he2016deep} has 152 layers.
The trend tend seem to go towards larger and deeper neural network.
The input layer of a neural network (in our example, a classifier) contains the (possibly pre-processed) inputs, such as images (for image recognition), words (in natural language processing), or motifs (in graph processing).
In our humble example, the input is an image of an apple, where each pixel has an all-to-all connection to the hidden layer-- a so-called \textit{dense} layer. The dense-layer (or fully-connected layer) is the most basic of all layers, and nearly all (e.g.\ convolution) layers are derived from it.
Each layers consists of a number of \textit{neurons} – the main compute agent of the network.
Each edge that connect to a particular neuron is called a \textit{synapse}, and comes with a \textit{weight} (\(w_{i,j}\)) in the figure, where \(i\) is the index of the neuron).
The neuron integrates the weights (\cref{fig:ann-example}:b) of all input synapses, adds a local bias, and applies a activation function to compute the output of the neuron. The activation function can varies between networks, but commonly used activation functions include: the sigmoid (\(f(x) = \frac{1}{1+e^{-x}}\)), the rectifier (ReLu, \(f(x) = max(0,x)\)), or hyperbolic tangent (\(f(x) = tanh(x)\)).
In \cref{fig:ann-example}:b (and commonly in image processing network), the simple and compute-\textit{in}expensive ReLu activation function is used in the hidden-layers of the network. The final layer of our classifier neural network is an output layer, \cref{fig:ann-example}:c. In the output layer, the activation function is often a \textit{softmax} (\cref{fig:ann-example}:d), which scales all outputs such that their sum equals one.

@@

@@

\section{Methodology}\label{sec:method}
\subsection{Experimental Platform}
The software versions and hardware configuration are shown in respectively \cref{tab:software-versions} and \cref{tab:test-hardware}.
All kernels were compiled with \texttt{-O3 -fp-relaxed}.
\texttt{-fp-relaxed} gives the compiler some more freedom with the order of operations, this gives about a \SIrange{10}{15}{\percent} performance benefit without changing the numerical output.
\begin{table}[h]
  \centering
  \caption{Used software versions}\label{tab:software-versions}
  \begin{tabular}{ll}
    \toprule
    \textbf{Software} & \textbf{Version} \\
    \midrule
    Python & v3.6.3 \\
    Keras & v2.2.0 \\
    TensorFlow & v1.8.0 \\
    Intel Quartus & v17.1.0 \\
    \bottomrule
  \end{tabular}
\end{table}
\begin{table}[h]
  \centering
  \caption{Test hardware}\label{tab:test-hardware}
  \begin{tabular}{ll}
    \toprule
    \textbf{Hardware} & \textbf{Configuration} \\
    \midrule
    Intel i7-3930K & \SI{3.20}{\giga\hertz} \\
    DDR3 RAM & \SI{1333}{\mega\hertz} Dual Channel \\
    Arria 10 ($\times$1) & Nallatech 510T \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Network Tested + Data-set}
The used example networks are a simple multi-layer perceptron.
So all layers are a simple array neurons and every neuron is connected to every neuron in the next layer.
These layers are often called ``Dense'' layers.
The first data set used was MNIST, MNIST is a collection of 28\(\times\)28 gray scale images of hand written arabic numerals.
So the data set provides \num{784} input features.
The structure of the sample MNIST network is.
\begin{enumerate}
  \item Dense(512)  \(\to\)  Relu
  \item Dense(512)  \(\to\)  Relu
  \item Dense(10)  \(\to\)  Softmax
\end{enumerate}
So in total a three layer network, or threes sets of weight matrices and bias vectors.
The data set was trained using the first \num{60000} images from the MNIST data set.
The last \num{10000} images are used as the test data set.
The second data set used was CIFAR10, so \num{3072} input features.
The structure of the sample CIFAR10 network is.
\begin{enumerate}
  \item Dense(1024)  \(\to\)  Relu
  \item Dense(512)  \(\to\)  Relu
  \item Dense(512)  \(\to\)  Relu
  \item Dense(10)  \(\to\)  Softmax
\end{enumerate}

@@

\begin{table}[h]
  \centering
  \caption{FPGA Kernel performance}\label{tab:kernel-performance}
  \begin{tabular}{llll}
    \toprule
    \textbf{Block size} & \textbf{GFLOPS MNIST}  & \textbf{GFLOPS CIFAR10}  \\
    \midrule
    64$\times$64 & \num{17.8} & \num{18.9}  \\ % kernel_nn_blocking_B64inorder_hw_2018-06-28T22.52.27.aocx Network MNIST: [28 * 28, 512, 512, 10] CIFAR: [32 * 32 * 3, 1024, 512, 512, 10]
    16$\times$16 & \num{4.3} & \num{4.4} \\ % kernel_nn_blocking_B16inorder_hw_2018-06-28T22.53.27.aocx Network MNIST: [28 * 28, 512, 512, 10] CIFAR: [32 * 32 * 3, 1024, 512, 512, 10]
    \bottomrule
  \end{tabular}
\end{table}

@@

\subsection{Training Performance}
The MNIST network accuracy after training on the CPU using Keras is \SI{97.9}{\percent}.
And the CIFAR10 accuracy after training is \SI{47.9}{\percent}
The FPGA gets slightly different floating point results due to rounding differences but is matches the accuracy.

@@

\bibitem{msbrainwave}
  Chung, Eric and Fowers, Jeremy and Ovtcharov, Kalin and Papamichael, Michael and Caulfield, Adrian and Massengill, Todd and Liu, Ming and Lo, Daniel and Alkalay, Shlomi and Haselman, Michael and others:
  Serving DNNs in Real Time at Datacenter Scale with Project Brainwave:
  {\it IEEE Micro},
  vol.
38, number.
2, pp 8--20, 2018

\bibitem{googletpu}
  Google Blog (online):
  Google Supercharges Machine Learning Tasks with TPU Custom Chip,
  {\it https://cloudplatform.googleblog.com/2016/05/Google-supercharges-machine-learning-tasks-with-custom-chip.html}
  Accessed: 2018-06-25

\bibitem{fujitsudlu}
  Fujitsu:
  Post-K Development and Introducing DLU - Fujitsu,
  {\it http://www.fujitsu.com/global/Images/post-k-development-and-introducing-dlu.pdf},
  2016

@@

\bibitem{volta100}
  NVIDIA:
  Artificial Intelligence Architecture,
  {\it https://www.nvidia.com/en-us/data-center/volta-gpu-architecture/},
  Accessed: 2018-06-25

@@

\bibitem{tpuformat}
  The Next Platform:
  Tearing Apart Google’s TPU 3.0 AI CoProcessor,
  {\it https://www.nextplatform.com/2018/05/10/tearing-apart-googles-tpu-3-0-ai-coprocessor},
  Accessed: 2018-06-25

@@

\bibitem{nervana}
  Intel:
  Intel Nervana Neural Network Processor,
  {\it https://ai.intel.com/intel-nervana-neural-network-processor/},
  Access: 2018-06

\bibitem{xilinxml}
  Xilinx:
  Xilinx ML Suite,
  {\it https://www.xilinx.com/applications/megatrends/machine-learning.html},
  Accessed: 2018-16

>>>>

  %\usepackage[dvips]{graphicx}
  \usepackage{latexsym}
  \usepackage[hyphens]{url}
  \Urlmuskip=0mu plus 1mu

@@

FPGAs can be used for a great variety of things, from accelerating computational algorithms to implementing a hardware design before taping out full chip.
They are devices that can represent reconfigurable hardware.
A normal FPGA work-flow consists of writing in a hardware description language.
Then synthesizing the design and finally placing and routing the design.
This is a lengthy process and complex designs in hardware description languages are hard to maintain.
In this work we will focus on using the OpenCL interface to the FPGA.
OpenCL is a framework specified by the Khronos Group\footnote{Specification available at: \url{https://www.khronos.org/opencl/}} in 2009, to work with all kinds of processors, accelerators and other heterogenous compute devices, in recent years the FPGA vendors have made efforts to make it easier to leverage FPGA resources by computer scientists and not only hardware designers.
So OpenCL is one of the ways to leverage high level synthesis.
High level synthesis take a design specification in a high level language like C or in this case the modified version of C99 that is used for OpenCL kernels.
And the synthesizer makes an RTL design description from this design automatically after which the normal work-flow is resumed.

FPGAs are made up of a reconfigurable fabric.
The fabric they are made up of consists of vast arrays of switches and other routing hardware to connect all the functional elements, like multipliers, block RAMs and LUTs, in any way possible.
The LUTs can be used to implement any mathematical function, like addition, logical function et cetera.
This makes them very suitable for the research described in this paper.
It gives total freedom to design a hardware implementation to explore algorithms in a hardware setting without having to resort to software simulations and the performance limitations that follow from that.

@@

\subsection{Neural Networks}
An Artificial Neural Network (ANNs) is a system inspired by the biological nervous system-- a system built around the concepts of \textit{neuron} as the main computational agents.
\Cref{fig:ann-example} illustrates an example of a typical “deep” neural network: a network that consists of a input layer, and output layer, and a number of \textit{hidden} layers in-between.
The number of hidden layers in modern neural networks can be very large: AlexNet (2012)~\cite{krizhevsky2012imagenet} has eight, VGG16 (2014)~\cite{simonyan2014very} has 16 layers, GoogleNet (2014)~\cite{szegedy2015going} has 22 layers, and ResNet (2016)~\cite{he2016deep} has 152 layers.
The trend tend seem to go towards larger and deeper neural networks.
The input layer of a neural network (in our example, a classifier) contains the (possibly pre-processed) inputs, such as images (for image recognition), words (in natural language processing), or motifs (in graph processing).
In our humble example, the input is an image of an apple, where each pixel has an all-to-all connection to the hidden layer-- a so-called \textit{dense} layer. The dense-layer (or fully-connected layer) is the most basic of all layers, and nearly all (e.g.\ convolution) layers are derived from it.
Each layers consists of a number of \textit{neurons} – the main compute agent of the network.
Each edge that connect to a particular neuron is called a \textit{synapse}, and comes with a \textit{weight} (\(w_{i,j}\)) in the figure, where \(i\) is the index of the neuron).
The neuron integrates the weights (\cref{fig:ann-example}:b) of all input synapses, adds a local bias, and applies a activation function to compute the output of the neuron. The activation function can varies between networks, but commonly used activation functions include: the sigmoid (\(f(x) = \frac{1}{1+e^{-x}}\)), the rectifier (ReLu, \(f(x) = max(0,x)\)), or hyperbolic tangent (\(f(x) = tanh(x)\)).
In \cref{fig:ann-example}:b (and commonly in image processing network), the simple and compute-\textit{in}expensive ReLu activation function is used in the hidden-layers of the network. The final layer of our classifier neural network is an output layer, \cref{fig:ann-example}:c. In the output layer, the activation function is often a \textit{softmax} (\cref{fig:ann-example}:d), which scales all outputs such that their sum equals one.

@@

The OpenCL SDK from Intel used for this design, tries to pipeline loops in single work item kernels, as opposed to NDRange kernels. % Reference to Intel OpenCL Manual?
To get maximum performance the aim is to get an initiation interval of one.
That is to say that a new loop iteration is scheduled every clock cycle.
To get this to work on the inner most loop, one that is essentially a reduction.
An array is used to store all output in separate locations, this way the loop can be fully unrolled.
For the containing loops, serial dependencies are to be avoided whenever possible.
So different loop iterations are fully data-independent.
For the outer most loop that processes one block this behavior is not possible, since the output block is written to multiple times.
If there was enough space on the FPGA theoretically multiple blocks would be able to be processed in parallel.
In the future the training kernel also needs to fit, so it was decided not to try parallelize this further.

@@

\section{Methodology}\label{sec:method}
\subsection{Experimental Platform}
The software versions and hardware configuration are shown in respectively \cref{tab:software-versions} and \cref{tab:test-hardware}.
All kernels were compiled with \texttt{-O3 -fp-relaxed}.
\texttt{-fp-relaxed} gives the compiler some more freedom with the order of operations, this gives about a \SIrange{10}{15}{\percent} performance benefit without changing the numerical output.
\begin{table}[h]
  \centering
  \caption{Used software versions}\label{tab:software-versions}
  \begin{tabular}{ll}
    \toprule
    \textbf{Software} & \textbf{Version} \\
    \midrule
    Python & v3.6.3 \\
    Keras & v2.2.0 \\
    TensorFlow & v1.8.0 \\
    Intel Quartus & v17.1.0 \\
    Intel OpenCL & v1.1 \\
    \bottomrule
  \end{tabular}
\end{table}
\begin{table}[h]
  \centering
  \caption{Test hardware}\label{tab:test-hardware}
  \begin{tabular}{ll}
    \toprule
    \textbf{Hardware} & \textbf{Configuration} \\
    \midrule
    Intel i7-3930K & \SI{3.20}{\giga\hertz} \\
    DDR3 RAM & \SI{1333}{\mega\hertz} Dual Channel \\
    Arria 10 ($\times$1) & Nallatech 510T \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Network Tested + Data-set}
The used example networks are a simple multi-layer perceptron.
So all layers are a simple array neurons and every neuron is connected to every neuron in the next layer.
These layers are often called ``Dense'' layers.
The first data set used was MNIST~\cite{lecun2010mnist}, MNIST is a collection of 28\(\times\)28 gray scale images of hand written arabic numerals.
So the data set provides \num{784} input features.
The structure of the sample MNIST network is.
\begin{enumerate}
  \item Dense(512)  \(\to\)  Relu
  \item Dense(512)  \(\to\)  Relu
  \item Dense(10)  \(\to\)  Softmax
\end{enumerate}
So in total a three layer network, or threes sets of weight matrices and bias vectors.
The data set was trained using the first \num{60000} images from the MNIST data set.
The last \num{10000} images are used as the test data set.

The second data set used was CIFAR-10~\cite{krizhevsky2014cifar}, so \num{3072} input features.
The structure of the sample CIFAR-10 network is.
\begin{enumerate}
  \item Dense(1024)  \(\to\)  Relu
  \item Dense(512)  \(\to\)  Relu
  \item Dense(512)  \(\to\)  Relu
  \item Dense(10)  \(\to\)  Softmax
\end{enumerate}

@@

\begin{table}[h]
  \centering
  \caption{FPGA Kernel performance}\label{tab:kernel-performance}
  \begin{tabular}{llll}
    \toprule
    \textbf{Block size} & \textbf{GFLOPS MNIST}  & \textbf{GFLOPS CIFAR-10}  \\
    \midrule
    64$\times$64 & \num{17.8} & \num{18.9}  \\ % kernel_nn_blocking_B64inorder_hw_2018-06-28T22.52.27.aocx Network MNIST: [28 * 28, 512, 512, 10] CIFAR: [32 * 32 * 3, 1024, 512, 512, 10]
    16$\times$16 & \num{4.3} & \num{4.4} \\ % kernel_nn_blocking_B16inorder_hw_2018-06-28T22.53.27.aocx Network MNIST: [28 * 28, 512, 512, 10] CIFAR: [32 * 32 * 3, 1024, 512, 512, 10]
    \bottomrule
  \end{tabular}
\end{table}

@@

\subsection{Training Performance}
The MNIST network test set accuracy after training on the CPU using Keras is \SI{97.9}{\percent}.
And the CIFAR-10 accuracy after training is \SI{47.9}{\percent}
The FPGA gets slightly different floating point results due to rounding differences but is matches the accuracy.

@@

\bibitem{lecun2010mnist}
  LeCun, Yann and Cortes, Corinna and Burges, CJ:
  MNIST handwritten digit database,
  {\it AT\&T Labs \url{http://yann.lecun.com/exdb/mnist}},
  Accessed: 2018-07, 2010

\bibitem{krizhevsky2014cifar}
  Krizhevsky, Alex, Vinod Nair, and Geoffrey Hinton:
  The CIFAR-10 dataset,
  {\it \url{http://www.cs.toronto.edu/kriz/cifar.html}},
  Accessed: 2018-07, 2014

\bibitem{msbrainwave}
  Chung, Eric and Fowers, Jeremy and Ovtcharov, Kalin and Papamichael, Michael and Caulfield, Adrian and Massengill, Todd and Liu, Ming and Lo, Daniel and Alkalay, Shlomi and Haselman, Michael and others:
  Serving DNNs in Real Time at Datacenter Scale with Project Brainwave,
  {\it IEEE Micro},
  vol. 38, number. 2, pp 8--20, 2018

\bibitem{googletpu}
  Google Blog (online):
  Google Supercharges Machine Learning Tasks with TPU Custom Chip,
  {\it \url{https://cloudplatform.googleblog.com/2016/05/Google-supercharges-machine-learning-tasks-with-custom-chip.html}},
  Accessed: 2018-06-25

\bibitem{fujitsudlu}
  Fujitsu:
  Post-K Development and Introducing DLU - Fujitsu,
  {\it \url{http://www.fujitsu.com/global/Images/post-k-development-and-introducing-dlu.pdf}},
  2016

@@

\bibitem{volta100}
  NVIDIA:
  Artificial Intelligence Architecture,
  {\it \url{https://www.nvidia.com/en-us/data-center/volta-gpu-architecture/}},
  Accessed: 2018-06-25

@@

\bibitem{tpuformat}
  The Next Platform:
  Tearing Apart Google’s TPU 3.0 AI CoProcessor,
  {\it \url{https://www.nextplatform.com/2018/05/10/tearing-apart-googles-tpu-3-0-ai-coprocessor}},
  Accessed: 2018-06-25

@@

\bibitem{nervana}
  Intel:
  Intel Nervana Neural Network Processor,
  {\it \url{https://ai.intel.com/intel-nervana-neural-network-processor/}},
  Access: 2018-06

\bibitem{xilinxml}
  Xilinx:
  Xilinx ML Suite,
  {\it \url{https://www.xilinx.com/applications/megatrends/machine-learning.html}},
  Accessed: 2018-16
